Legion: Expressing Locality and Independence with Logical Regions

### 2012-05-04 doc/sc2012/experiments.tex
Modifications to the experiments section

In contrast, the machine model used by the Legion runtime is based on an
affinity graph of CPUs and GPUs and is able to accurately capture complex
machine hierarchies.

For each application, multiple problem sizes were used, and each size problem was
run on subsets of each machine ranging from the smallest (a single CPU core or GPU)
to the largest or near-largest (except Keeneland which has 115 nodes, but we limited 
our runs to at most 32 nodes to get sufficient cluster time.)
By looking at the performance of the same size problem over progressively larger
machines, we are able to measure the strong scaling that Legion is able to
archive.  By increasing the problem size as well, we also measure weak scaling.

@@

The first experiment we investigate is the distributed circuit simulation described in 
Section~\ref{sec:ex}.  The Legion runtime handles all of the resource allocation, 
scheduling, and data movement across the cluster of GPUs.  In particular,  
Legion's ability to efficiently move the irregularly partitioned
shared data around the system while keeping the private nodes and wires resident in
each GPU's framebuffer memory is critical in keeping the overhead low and achieving 
good scalability.

@@

Figure~\ref{fig:ckt_overhead} shows the fraction of the overall simulation time (summed over
all nodes) spent in the application kernels compared to the various pieces of the Legion
runtime.  As the node count increases, the non-communication overhead stays relatively constant.
As expected the communication overhead grow linearly with number of nodes.

@@

Our second experiment is a port of the \emph{fluidanimate} benchmark from the PARSEC suite\cite{bienia11benchmarking},
which does a particle-based simulation of an incompressible fluid.  Each particle interacts only with other nearby
particles. The benchmark divides the space in which the fluid can move into a three-dimensional array of cells 
such that the range of interaction is limited to just the
cells adjacent (including diagonals) to the one a particle resides in.  Originally designed to be a
chip multiprocessor benchmark, the application divides the array into {\em grids} and assigns each one to 
a thread.  Per cell locking is used to safely access particles in cells that lie
on the edge of a grid.  This fine-grained locking scheme along with the assumption of a shared address space lead to
good scaling in a multi-core processor, but rule out any attempt run beyond a single node.  To extend the
scaling further, our port uses region partitioning to divide the array into grids in a similar
way, but handles the interaction between grids in a way that doesn't rely on access to uniform shared memory.
Instead, the Legion version of the benchmark creates explicit ghost copies of cells on the boundaries of grids
and uses those ghost cells to exchange information between the grids.

@@

Figure~\ref{fig:fluid_single} compares the performance of the Legion implementation against the PARSEC version,
using a relatively small problem (300K particles on a 15x21x15 array of cells).  Between 1 and 4 threads, the
results for sapling are nearly indistinguishable, indicating 
neither the Legion runtime nor the the restructuring of the implementation to allow multi-node scaling impose
any significant overhead.  It's possible that the use of explicit ghost cells rather than 
fine-grained sharing of cache lines might be a net win.  At 8 threads and above, the performance begins to vary.
Both the Legion and PARSEC version on viz flatten out as they over-subscribe the 12 physical cores.  On sapling,
which has HyperThreading enabled, deviations from linear begin sooner as the operating system's scheduler's
thread placement choices begin to matter.

To measure scaling beyond a single node, three different problem sizes were run for each of the three systems,
 with the results plotted in Figure~\ref{fig:fluid_multi}.  
 For the smallest problem (300K particles), we observe a 20\% speedup when going to 1
node to 2 (a total of 16 threads), but begin slowing down beyond that due to communication overhead - at 4 nodes
there are twice as many ghost cells as interior grid cells.  The larger problem sizes (2.4M and 19M
particles) do much better, with scaling of up to 5.4x when going from 1 node up to 16 because the
communication-to-computation ratio is lower.

@@

\subsection{Adaptive Mesh Refinement}
\label{subsec:exp_amr}
Our final application is based on the third heat equation example from the Berkeley Labs 
BoxLib project \cite{BoxLib}.  This application is a three-level adaptive-mesh-refinement (AMR)
code that computes a first order stencil on a 2D mesh of cells.  The simulation iterates for many time
steps.  Updating the simulation for one time step consists of three phases.  In the first phase,
the boundary cells around a box at a refined level linearly interpolate their values from the nearby cells at the
next coarsest level.  The second phase computes the stencil computation on each cell in all levels.
In the third phase, cells at a coarser level that have been refined are restricted to the average
of the cells that they physically contain at the next finest level of refinement.

@@

Our second optimization takes advantage of the cross-level locality present in the application.  We implemented
an application specific mapper that inspects the arguments of tasks to discover the relationships between
grids at different levels of refinement.  The mapper dynamically performs intersection tests between logical regions
containing grids of different refinement levels to determine whether they overlap.  
If grids from different levels overlap, the mapper
ensures that they are always placed on the same node in the machine to leverage cross-level locality.  The mapper
memoizes the results of these intersection tests to amortize their cost.  The mapper also dynamically
load balances by distributing unconstrained grids from the coarsest level onto under-loaded nodes.

We compared our Legion implementation against BoxLib on three different problem sizes with a fixed number
of cells per level of refinement, but with randomly chosen refinements locations.  BoxLib also supports OpenMP
and we took their best number from using 1, 2, 4, or 8 OpenMP threads per node.  Our Legion implementation
always uses one thread per node to illustrate that in this application locality is significantly more important 
than parallelism.  The results from our experiments on all three clusters can be seen in Figure~\ref{fig:amr_total}.

@@

As the node count increased the AMR code becomes highly dependent on interconnect performance.  BoxLib performs much better
on Keeneland than it does on our viz cluster due to the better interconnect.  This can also be seen as we scale
to a larger number of nodes and BoxLib begins to catch up as can be seen in Figure~\ref{fig:amr_keeneland}.  
This occurs because our intra-level ghost-cell exchange algorithm relies on GASNet memory to exchange ghost cells which 
involves a linear increase in network traffic with the number of nodes.  BoxLib instead uses direct
node-to-node exchanges of ghost cells similar to our fluid application.  A future implementation of our AMR code 
will employ a similar ghost cell exchange algorithm to improve scalability.

>>>>

In contrast, the machine model used by the Legion SOOP runtime is based on an
affinity graph of CPUs and GPUs and is able to accurately capture complex
machine hierarchies.

For each application, multiple problem sizes were used, and each size problem was
run on subsets of each machine ranging from the smallest (a single CPU core or GPU)
to the largest or near-largest (except Keeneland which has 120 nodes, but we limited 
to at most 32 nodes to get sufficient cluster time.)
By looking at the performance of the same size problem over progressively larger
machines, we are able to measure the strong scaling that Legion is able to
archive.  By increasing the problem size as well, we also measure weak scaling.

@@

The first experiment we investigate is the distributed circuit simulation described in 
Section~\ref{sec:ex}.  The Legion SOOP runtime handles all of the resource allocation, 
scheduling, and data movement across the cluster of GPUs.  In particular,  
Legion's ability to efficiently move the irregularly partitioned
shared data around the system while keeping the private nodes and wires resident in
each GPU's framebuffer memory is critical to achieving good scalability.

@@

Figure~\ref{fig:ckt_overhead} shows the fraction of the overall simulation time (summed over
all nodes) spent in the application kernels compared to the various pieces of the Legion
SOOP runtime.  As the node count increases, the non-communication overhead stays relatively constant.
As expected the communication overhead grow linearly with number of nodes.

@@

Our second experiment is a port of the \emph{fluidanimate} benchmark from the PARSEC suite\cite{bienia11benchmarking},
which does a particle-based simulation of an incompressible fluid.  Each particle interacts only with other nearby
particles. The benchmark divides the space in which the fluid can move into a three-dimensional array of cells 
such that the range of interaction is limited to just the
cells adjacent (including diagonals) to the one a particle resides in.  The application divides the array 
into {\em grids} and assigns each one to 
a thread.  Per cell locking is used to safely access particles in cells that lie
on the edge of a grid.  This fine-grained locking scheme along with the assumption of a shared address space leads to
good scaling in a multi-core processor, but rules out any attempt run beyond a single node.

To extend the scaling further, our port uses region partitioning to divide the array into grids in a similar
way, but handles the interaction between grids in a way that doesn't rely on access to shared memory.
Instead, the Legion version of the benchmark creates explicit ghost copies of cells on the boundaries of grids
and uses those ghost cells to exchange information between the grids.

@@

Figure~\ref{fig:fluid_single} compares the performance of the Legion implementation against the PARSEC version,
using a relatively small problem (300K particles on a 15x21x15 array of cells).  Between 1 and 4 threads, the
results for Sapling are nearly indistinguishable, indicating 
neither the Legion runtime nor the restructuring of the implementation to allow multi-node scaling impose
any significant overhead.  
%It's possible that the use of explicit ghost cells rather than fine-grained sharing of cache lines might be a net win.  
At 8 threads and above, the performance begins to vary.
Both the Legion and PARSEC version on Viz flatten out as they over-subscribe the 12 physical cores.  On Sapling,
which has HyperThreading enabled, deviations from linear begin sooner as the operating system's scheduler's
thread placement choices begin to matter.

To measure scaling beyond a single node, three different problem sizes were run for each of the three systems,
 with the results plotted in Figure~\ref{fig:fluid_multi}.  
 For the smallest problem (300K particles), we observe a 20\% speedup when going to 1
node to 2 (a total of 16 threads), but begin slowing down beyond that due to communication overhead - at 4 nodes
there are twice as many ghost cells as interior grid cells.  The larger problem sizes (2.4M and 19M
particles) do much better, with scaling of up to 5.4x when going from 1 node up to 16 because of a lower 
communication-to-computation ratio.

@@

\subsection{Adaptive Mesh Refinement}
\label{subsec:exp_amr}
Our final application is based on the third heat equation example from the Berkeley Labs 
BoxLib project \cite{BoxLib}.  This application is a three-level adaptive-mesh-refinement (AMR)
code that computes a first order stencil on a 2D mesh of cells.  The simulation iterates for many time
steps.  Updating the simulation for one time step consists of three phases.  In the first phase,
the boundary cells around a box at a refined level linearly interpolate their values from the nearby cells at the
next coarsest level.  The second phase performs the stencil computation on each cell in every level.
In the third phase, cells at a coarser level that have been refined are restricted to the average
of the cells that they physically contain at the next finest level of refinement.

@@

Our second optimization takes advantage of the cross-level locality present in the application.  We implemented
an application specific mapper that dynamically discovers the relationships between
grids at different levels of refinement.  The mapper dynamically performs intersection tests between logical regions
containing grids of different refinement levels to determine whether they overlap.  
If grids from different levels overlap, the mapper
ensures that they are always placed on the same node in the machine to leverage cross-level locality.  The mapper
memoizes the results of these intersection tests to amortize their cost.  The mapper also dynamically
load balances by distributing unconstrained grids from the coarsest level onto under-loaded nodes.

We compared our Legion implementation against BoxLib on three different problem sizes with a fixed number
of cells per level of refinement, but with randomly chosen refinements locations.  BoxLib also supports OpenMP
and we took their best performance from using 1, 2, 4, or 8 threads per node.  Our Legion implementation
always uses one thread per node to illustrate that in this application locality is significantly more important 
than parallelism.  The results from our experiments on all three clusters can be seen in Figure~\ref{fig:amr_total}.

@@

As the node count increased the AMR code becomes highly dependent on interconnect performance.  BoxLib performs much better
on Keeneland than it does on our Viz cluster due to the better interconnect.  As we scale
to a larger number of nodes BoxLib begins to catch up as can be seen in Figure~\ref{fig:amr_keeneland}.  
This occurs because our intra-level ghost-cell exchange algorithm relies on GASNet memory to exchange ghost cells which 
involves a linear increase in network traffic with the number of nodes.  BoxLib instead uses direct
node-to-node exchanges of ghost cells similar to our fluid application.  A future implementation of our AMR code 
will employ a similar ghost cell exchange algorithm to improve scalability.
